\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-31 doc/publications/plos16/plos16.tex
fix typos

%% Additional packages
\usepackage{footmisc}
\usepackage{multirow}
\usepackage[T1]{fontenc}
%% required the installation of cm-super 55MB - TeX font package (full version) with CM (EC) in Type1 in T1, T2*, TS1, X2 enc

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
trees and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}).

\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
Check this. }

@@

\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interaction of large groups of SNPs.

@@

\subsubsection{Bone mineral density dataset}
We obtained a VCF file from the authors of \cite{Duncan.2011}, which contains variant data from 1,936 postmenopausal women.
This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan.2011}.

@@

Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
value for a classification problem is $\sqrt{p} \approx 1581$ so in this case we are considering values considerably larger than
the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largest
values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure depends
heavily on \ntree but gives a better valuye with lower values of \mtry.

@@

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/oob_levleplot.png}
      \caption{OOB classification error.} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo_levleplot.png}
      \caption{The rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
      the better result, that is, a lower OOB or a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
specific phenotype.  In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a
model on the BMD dataset (see datasets) using a binary high or low BMD value as the label.  From this model, we output a
list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity,
and hence importance, is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.  Then, using the 26
known and 6 novel BMD genes from \cite{Duncan.2011}, we compile a list of variants that are present in or near one of
these 32 genes.

@@

However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist
within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
contributing to BMD.

@@

We also frequently see highly-ranked variants in chromosome 22q11. \cite{Stagi.2010}, observe that individuals with a
microdeletion in this region have a significant in bone mass \cite{Stagi.2010}. In this same region, we also observe
variants present in PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as
proposed by \cite{Phang.2008}, proline is made available by collagen degradation \cite{Phang.2008}.

DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that \cite{Duncan.2011}.
failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurrence of a variant in
or near this gene has a rank of 9,667 out of 10,000. Also, this gene is not highly expressed in bone, and any mechanisms
by which this gene might regulate BMD are unclear \cite{Thakker2012}. Although variants near DCDC5 are associated with
BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.

@@

In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering \cite{Obrien2016}.
\marginpar{reference is missing}
Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
process (pre-processing and clustering) taking over 30 hours.
We investigate building random forest models on this data. However, we observed this process to not only be slower
than building a k-means model, but does not scale to as many dimensions (see Table~\ref{timingtable}). To keep
the following comparisons fair, we limit the number of executors available to Spark. On the relatively low-dimensional
phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger, 
presumably due to the overhead of Spark's distributed platform on this relatively small dataset, we observe Spark ML to
be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
in Spark ML's runtime to over 8 hours. Moving to our next largest dataset, phase3\_chr1\-3 (2,504 x 19,328,051), and the
job fails to complete within our arbitrary cutoff time of 20 hours.

From our trials, it was apparent that Spark ML, although designed to handle many samples, is unable to efficiently cope
with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
common.

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\centering
\caption{ {\bf Performance comparison with \cursedforest on different subsets of the 1000 Genomes Project data.}}
\begin{tabular}{| l | r | l |}
\hline
{\bf Genome \%} & {\bf Error} & {\bf Runtime} \\ \thickhline
phase1\_chr22: 1,092 x 490,036 & 0.04 & 1 min 5 seconds \\ \hline
phase3\_chr22: 2,504 x 1,103,548 & 0.06 & 1 min 37 seconds \\ \hline
phase3\_chr1: 2,504 x 6,450,364 & 0.04 & 4 min 10 seconds \\ \hline
phase3\_chr1-3: 2,504 x 19,328,051 & 0.02 & 10 min 49 seconds \\ \hline
phase3\_chr1-22: 2,504 x 81,047,467 & 0.01 & 36 min 54 seconds \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of runtime and error for models built using \cursedforest on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees and was built using 130 worker nodes, each with 8GB RAM.
\end{flushleft}
\label{cursedforesttable}
\end{minipage}
\end{table}

@@

We investigate the different synthetic datasets generated for section \ref{section:synthetic_data} and measured the time
taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases
were executed with the same random seed, to improve the consistency of measurements.

@@

\marginpar{tables 2 has been commented out -- there is no data in it. Piotr could add the data or we could remove both
  tables just leave the figures. I have done this for the time being} 
Next we investigate \cursedforest scalability with
regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
execution of 128 CPU cores. The results are visualized in Fig~\ref{figure:synth2} below (please note log scale on the
axes and the values on y axes are expresses as trees per hour).

Generally, the number of trees per hour decreases with increased number of variables and samples sizes. Some
irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that
keeping the \mtry fraction constant results in higher \mtry values with the growing number of variables, and this is what
drives the performance down rather that the increase of dataset size itself.

>>>>

%% Additional packages
\usepackage{footmisc}
\usepackage{multirow}
\usepackage[T1]{fontenc}

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
trees and use a high \mtry value which leads to greater accuracy (see the example in section \ref{section:synthetic}). 
\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
We do not see that. }

@@

Biau~{\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interaction of large groups of SNPs.

@@

\subsubsection{Bone mineral density dataset}
We obtained a VCF file from the authors of \cite{Duncan2011}, which contains variant data from 1,936 postmenopausal women.
This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan2011}.

@@

\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
sensitive once we over a certain threshold.

@@

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
      \caption{OOB classification error.} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo.png}
      \caption{The rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
      the better result, that is, a lower OOB or a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
specific phenotype.
In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a model on the BMD dataset
(see datasets) using a binary high or low BMD value as the label.
From this model, we output a list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with
the highest Gini impurity---and hence, importance---is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.
Then, using the 26 known and 6 novel BMD genes from Duncan~{\it et al.}, we compile a list of variants that are present in or near one of these
32 genes.

@@

However, we also observe that from the set of highest-ranked features---for example, the top ten---not all of them exist
within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
contributing to BMD.

@@

We also frequently see highly-ranked variants in chromosome 22q11. Stagi~{\it et al.} observe that individuals with a microdeletion
in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe variants present in
PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as proposed by Phang~{\it et al.}, proline
is made available by collagen degradation \cite{Phang2008}.

DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that Duncan~{\it et al.}.
failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurrence of a variant in
or near this gene has a rank of 9,667 out of 10,000. Also, this gene is not highly expressed in bone, and any mechanisms
by which this gene might regulate BMD are unclear \cite{Thakker2012}. Although variants near DCDC5 are associated with
BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.

@@

In a previous paper, we clustered individuals from the 1000 Genomes Project using k-means clustering\cite{Obrien2016}.
Although we were able to cluster individuals on all 80 million variants, it was relatively slow, with the entire
process (pre-processing and clustering) taking over 30 hours.
We investigate building random forest models on this data---however we observed this process to not only be slower
than building a k-means model, but also unable to scale to as many dimensions (see Table~\ref{timingtable}). To keep
the following comparisons fair, we limit the number of executors available to Spark. On the relatively low-dimensional
phase1\_chr22 dataset (1,092 x 490,036), we observed a runtime of under 8 minutes. Although slower than Ranger---
presumably due to the overhead of Spark's distributed platform on this relatively small dataset---we observe Spark ML to
be about 16 minutes faster than R. Moving to the larger dataset, phase3\_chr1 (2,504 x 6,450,364), we observe an increase
in Spark ML's runtime to over 8 hours. Moving to our next largest dataset, phase3\_chr1\-3 (2,504 x 19,328,051), and the
job fails to complete within our arbitrary cutoff time of 20 hours.

From our trials, it was apparent that Spark ML---although designed to handle many samples---is unable to efficiently cope
with a huge number of features. With the decreasing costs of genome sequencing, this case will surely become increasingly
common.

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\centering
\caption{ {\bf Performance comparison with \cursedforest on different subsets of the 1000 Genomes Project data.}}
\begin{tabular}{| l | r | l |}
\hline
{\bf Genome \%} & {\bf Error} & {\bf Runtime} \\ \thickhline
phase1\_chr22: 1,092 x 490,036 & 0.04 & 1 min 5 seconds \\ \hline
phase3\_chr22: 2,504 x 1,103,548 & 0.06 & 1 min 37 seconds \\ \hline
phase3\_chr1: 2,504 x 6,450,364 & 0.04 & 4 min 10 seconds \\ \hline
phase3\_chr1-3: 2,504 x 19,328,051 & 0.02 & 10 min 49 seconds \\ \hline
phase3\_chr1-22: 2,504 x 81,047,467 & 0.01 & 36 min 54 seconds \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of runtime and error for models built using \cursedforest on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees and was built using 130 worker nodes---each with 8GB RAM.
\end{flushleft}
\label{cursedforesttable}
\end{minipage}
\end{table}

@@

We investigate the different synthetic datasets generated for section \ref{ection:synthetic_data} and measured the time
taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases
were executed with the same random seed, to improve the consistency of measurements.

@@

\marginpar{tables 2 has been commented out -- there is no data in it. Piotr could add the data or we could remove both
  tables just leave the figures. I have done this for the time being} 
Next we investigate \cursedforest scalability with
regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
execution of 128 CPU cores. The results are visualized in Fig~\ref{figure:synth2} below (please note log scale on the
axes and the values on y axes are expressed as trees per hour).

Generally, the number of trees per hour decreases with an increased number of variables and samples sizes. Some
irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that
keeping the \mtry fraction constant results in higher \mtry values with the growing number of variables, and this is what
drives the performance down rather that the increase of dataset size itself.
